---
title: "p8105_hw2_qh2284"
author: "Qin Huang"
date: "2023-09-26"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidyr)
library(lubridate)
library(dplyr)
```

# Problem 1

**First, clean the data in pols-month.csv. Use `separate()` to break up the variable `mon` into integer variables `year`, `month`, and `day`; replace month number with month name; create a `president` variable taking values `gop` and `dem`, and remove `prez_dem` and `prez_gop`; and remove the day variable.**

```{r}
# import the dataset
pols = read_csv("pols-month.csv")

# clean the data
pols = janitor::clean_names(pols)

```

```{r}
# break up the date
pols = separate(pols, mon, into = c("year", "month", "day"), sep = "-")

# replace month number with month name
pols$month = month.name[as.numeric(pols$month)]

# create a president variable taking values gop and dem, and remove prez_dem and prez_gop
pols = pols %>%
  mutate(president = if_else(prez_dem == 1,'dem','gop')) %>%
  # remove prez_dem and prez_gop, day
  dplyr::select(-prez_dem, -prez_gop, -day) 
```

**Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that `year` and `month` are the leading columns.**

```{r}
# import the dataset
snp = read_csv("snp.csv")

# clean the data
snp = janitor::clean_names(snp)
```

```{r}
# separate year and month
snp = separate(snp, date, into = c("month", "day", "year"), sep = "/")

# convert 2-digit year to 4-digit year
snp = snp %>%
  mutate(year = if_else(year >= 0 & year <= 15, paste("20", year, sep = ""),paste("19", year, sep = ""))) %>%
  dplyr::select(year, month,everything()) %>%
  dplyr::select(-day)

# arrange according year and month
snp = snp %>% arrange(year, as.numeric(month))

# convert month number to name
snp$month = month.name[as.numeric(snp$month)]

```

**Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from \"wide\" to \"long\" format; ensuring that key variables have the same name; and ensuring that key variables take the same values.**

```{r}
# import the dataset
une = read_csv("unemployment.csv")
```

```{r}
# convert table from wide to long
une_new = une %>%
  pivot_longer(
    Jan:Dec,
    names_to = "month",
    values_to = "percentage"
  )

# convert month from abb to name
une_new$month = month.name[match(une_new$month, month.abb)]

# clean the data
une_new = janitor::clean_names(une_new)
```

**Join the datasets by merging `snp` into `pols`, and merging `unemployment` into the result.**

```{r}
# create a key "date"
pols_date = pols %>% 
  mutate(date = paste0(year, " ", month), .before = 1) %>%
  dplyr::select(-year,-month)
  
snp_date = snp %>% 
  mutate(date = paste0(year, " ", month), .before = 1) %>%
  dplyr::select(-year,-month)

une_date = une_new %>%
  mutate(date = paste0(year, " ", month), .before = 1) %>%
  dplyr::select(-year,-month)

# join the datasets by merging snp into pols
pols_add_snp = inner_join(pols_date, snp_date, by = 'date')

merge_table = inner_join(pols_add_snp, une_date, by = 'date')
```

**Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).**

1.  `pols-month.csv` dataset has 822 rows (observations) and 9 columns (variables). The range of the year (month) is from 1947.01 to 2015.06. The key variables are `mon`, `prez_gop`, `gov_gop`, `sen_gop`, `rep_gop`, `prez_dem`, `gov_dem`, `sen_dem`, `rep_dem`.

    The resulting dataset `pols` has 822 rows (observations) and 9 columns (variables). The range of the year (month) is from 1947 January to 2015 June. The key variables are `year`, `month`, `gov_gop`, `sen_gop`, `rep_gop`, `gov_dem`, `sen_dem`, `rep_dem`, `president`.

2.  `snp.csv` dataset has 787 rows (observations) and 2 columns (variables). The range of the year (month) is from 1950.01 to 2015.07. The key variables are `date`, `close`.

    The resulting dataset `snp` has 787 rows (observations) and 3 columns (variables). The range of the year (month) is from 1950 January to 2015 July. The key variables are `year`, `month`, `close`.

3.  `unemployment.csv` dataset has 68 rows (observations) and 13 columns (variables). The range of the year (month) is from 1948 Jan to 2015 Jun. The key variables are `Year`, `Jan`, `Feb`, `Mar`, `Apr`, `May`, `Jun`, `Jul`, `Aug`, `Sep`, `Oct`, `Nov`, `Dec`.

    The resulting dataset `une_new` has 816 rows (observations) and 3 columns (variables). The range of the year (month) is from 1948 January to 2015 June. The key variables are `year`, `month`, `percentage`.

------------------------------------------------------------------------

# Problem 2
