---
title: "p8105_hw2_qh2284"
author: "Qin Huang"
date: "2023-09-26"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidyr)
library(lubridate)
library(dplyr)
```

# Problem 1

**First, clean the data in pols-month.csv. Use `separate()` to break up the variable `mon` into integer variables `year`, `month`, and `day`; replace month number with month name; create a `president` variable taking values `gop` and `dem`, and remove `prez_dem` and `prez_gop`; and remove the day variable.**

```{r}
# import the dataset
pols = read_csv("./dataset/pols-month.csv")

# clean the data
pols = janitor::clean_names(pols)

```

```{r}
# break up the date
pols = separate(pols, mon, into = c("year", "month", "day"), sep = "-")

# create a president variable taking values gop and dem, and remove prez_dem and prez_gop
pols = pols %>%
  # replace month number with month name
  mutate(month = month.name[as.numeric(month)]) %>%
  mutate(president = if_else(prez_dem == 1,'dem','gop')) %>%
  # remove prez_dem and prez_gop, day
  dplyr::select(-prez_dem, -prez_gop, -day) 
```

**Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that `year` and `month` are the leading columns.**

```{r}
# import the dataset
snp = read_csv("./dataset/snp.csv")

# clean the data
snp = janitor::clean_names(snp)
```

```{r}
# separate year and month
snp = separate(snp, date, into = c("month", "day", "year"), sep = "/")

# convert 2-digit year to 4-digit year
snp = snp %>%
  mutate(year = if_else(year >= 0 & year <= 15, paste("20", year, sep = ""),paste("19", year, sep = ""))) %>%
  dplyr::select(year, month,everything()) %>%
  dplyr::select(-day)

# arrange according year and month
snp = snp %>% arrange(year, as.numeric(month))

# convert month number to name
snp = snp %>%
  mutate(month = month.name[as.numeric(month)])

```

**Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from "wide" to "long" format; ensuring that key variables have the same name; and ensuring that key variables take the same values.**

```{r}
# import the dataset
une = read_csv("./dataset/unemployment.csv")
```

```{r}
# convert table from wide to long
une_new = une %>%
  pivot_longer(
    Jan:Dec,
    names_to = "month",
    values_to = "percentage"
  )

# convert month from abb to name
une_new = une_new %>%
  mutate(month = month.name[match(month, month.abb)])

# clean the data
une_new = janitor::clean_names(une_new)
```

**Join the datasets by merging `snp` into `pols`, and merging `unemployment` into the result.**

```{r}
# create a key "date"
pols_date = pols %>% 
  mutate(date = paste0(year, " ", month), .before = 1) %>%
  dplyr::select(-year,-month)
  
snp_date = snp %>% 
  mutate(date = paste0(year, " ", month), .before = 1) %>%
  dplyr::select(-year,-month)

une_date = une_new %>%
  mutate(date = paste0(year, " ", month), .before = 1) %>%
  dplyr::select(-year,-month)

# join the datasets by merging snp into pols
pols_add_snp = left_join(pols_date, snp_date, by = 'date')

merge_table = left_join(pols_add_snp, une_date, by = 'date')
```

**Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).**

1.  `pols-month.csv` dataset has `r nrow(pols)` rows (observations) and `r ncol(pols)` columns (variables). The range of the year (month) is from 1947.01 to 2015.06. The key variables are `mon`, `prez_gop`, `gov_gop`, `sen_gop`, `rep_gop`, `prez_dem`, `gov_dem`, `sen_dem`, `rep_dem`.

    The resulting dataset `pols` has `r nrow(pols)` rows (observations) and `r ncol(pols)` columns (variables). The range of the year (month) is from 1947 January to 2015 June. The key variables are `year`, `month`, `gov_gop`, `sen_gop`, `rep_gop`, `gov_dem`, `sen_dem`, `rep_dem`, `president`.

2.  `snp.csv` dataset has `r nrow(snp)` rows (observations) and 2 columns (variables). The range of the year (month) is from 1950.01 to 2015.07. The key variables are `date`, `close`.

    The resulting dataset `snp` has `r nrow(snp)` rows (observations) and 3 columns (variables). The range of the year (month) is from 1950 January to 2015 July. The key variables are `year`, `month`, `close`.

3.  `unemployment.csv` dataset has `r nrow(une)` rows (observations) and `r ncol(une)` columns (variables). The range of the year (month) is from 1948 Jan to 2015 Jun. The key variables are `Year`, `Jan`, `Feb`, `Mar`, `Apr`, `May`, `Jun`, `Jul`, `Aug`, `Sep`, `Oct`, `Nov`, `Dec`.

    The resulting dataset `une_new` has `r nrow(une_new)` rows (observations) and `r ncol(une_new)` columns (variables). The range of the year (month) is from 1948 January to 2015 June. The key variables are `year`, `month`, `percentage`.

------------------------------------------------------------------------

# Problem 2

**Read and clean the Mr. Trash Wheel sheet:**

-   **specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in `read_excel`**

-   **use reasonable variable names**

-   **omit rows that do not include dumpster-specific data**

```{r}
# load the dataset, choose the range to skip the figure in the first row
library(readxl)
MrTrash = read_excel("./dataset/202309 Trash Wheel Collection Data.xlsx", sheet = 'Mr. Trash Wheel', range = "A2:N586")
```

```{r}
# use reasonalble variable names
MrTrash = janitor::clean_names(MrTrash)

MrTrash = separate(MrTrash, date, into = c("Year", "mon", "day"), sep = "-")

MrTrash = MrTrash %>% 
  dplyr::select(-Year, -mon) %>%
  dplyr::select(dumpster, year, month, everything())
```

**The data include a column for the (approximate) number of homes powered. This calculation is described in the `Homes powered note`, but not applied to every row in the dataset. Update the data to include a new `homes_powered` variable based on this calculation.**

```{r}
# update the home powered
MrTrash = MrTrash %>%
  mutate(homes_powered = weight_tons * 500 / 30)
```

**Use a similar process to import, clean, and organize the data for Professor Trash Wheel and Gwynnda.**

```{r}
# load the dataset, choose the range to skip the figure in the first row
Professor = read_excel("./dataset/202309 Trash Wheel Collection Data.xlsx", sheet = 'Professor Trash Wheel', range = "A2:M108")

# use reasonalble variable names
Professor = janitor::clean_names(Professor)

Professor = separate(Professor, date, into = c("Year", "mon", "day"), sep = "-")

Professor = Professor %>% 
  dplyr::select(-Year, -mon) %>%
  dplyr::select(dumpster, year, month, everything())

# update the home powered
Professor = Professor %>%
  mutate(homes_powered = weight_tons * 500 / 30)
```

```{r}
# load the dataset, choose the range to skip the figure in the first row
Gwynnda = read_excel("./dataset/202309 Trash Wheel Collection Data.xlsx", sheet = 'Gwynnda Trash Wheel', range = "A2:L157")

# use reasonalble variable names
Gwynnda = janitor::clean_names(Gwynnda)

Gwynnda = separate(Gwynnda, date, into = c("Year", "mon", "day"), sep = "-")

Gwynnda = Gwynnda %>% 
  dplyr::select(-Year, -mon) %>%
  dplyr::select(dumpster, year, month, everything())

# update the home powered
Gwynnda = Gwynnda %>%
  mutate(homes_powered = weight_tons * 500 / 30)
```

**Combine these with the Mr. Trash Wheel dataset to produce a single tidy dataset. To keep track of which Trash Wheel is which, you may need to add an additional variable to all datasets before combining.**

```{r}
# add a column to each dataset for merging
MrTrash = MrTrash %>%
  mutate(source = "Mr. Trash") %>%
  mutate(year = as.character(year))

Professor = Professor %>%
  mutate(source = "Professor Trash") %>%
  mutate(year = as.character(year))

Gwynnda = Gwynnda %>%
  mutate(source = "Gwynnda Trash") %>%
  mutate(year = as.character(year))

# combine data
total_trash = bind_rows(MrTrash, Professor, Gwynnda)

# ordered by the date
total_trash_date = total_trash %>%
  arrange(year, match(month, month.name), day)
```

**Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables.**

MrTrash has `r nrow(MrTrash)` observations and `r ncol(MrTrash)` variables. The dataset contains the weight of different kinds of trash.

Professor has `r nrow(Professor)` observations and `r ncol(Professor)` variables. The dataset contains the weight of different kinds of trash.

Gwynnda has `r nrow(Gwynnda)` observations and `r ncol(Gwynnda)` variables. The dataset contains the weight of different kinds of trash.

The number of observations in the resulting dataset is `r nrow(total_trash)` . The key variables such as `year`, `month`, `day`, represent the record time. `dumpster` represents the "id". `weight` is the variable that represents the weight of the trash which we used to calculate the home being powered. `home_powered`, as mentioned before, is the variable that represents the amount of the home being powered by the trash. `source` represents which sheet the data comes from. Other variables focus on different trash being collected.

**For available data, what was the total weight of trash collected by Professor Trash Wheel?**

```{r}
sum_weight_pro = sum(pull(Professor, weight_tons))
```

The total weight of trash collected by Professor Trash Wheel is `r sum_weight_pro` tons.

**What was the total number of cigarette butts collected by Gwynnda in July of 2021?**

```{r}
filtered_Gwynnda = Gwynnda %>%
  filter(year == 2021 & month == "July")

sum_ciga_Gwy = sum(pull(filtered_Gwynnda, cigarette_butts))
```

The total number of cigarette butts collected by Gwynnda in July of 2021 is `r sum_ciga_Gwy`.

------------------------------------------------------------------------

# Problem 3

**Import, clean, and tidy the dataset of baseline demographics. Ensure that sex and APOE4 carrier status are appropriate encoded (i.e. not numeric), and remove any participants who do not meet the stated inclusion criteria (i.e. no MCI at baseline).**

```{r}
# load the basline file
mci_baseline = read.csv('./dataset/MCI_baseline.csv', skip = 1)

# clean the data
mci_baseline = janitor::clean_names(mci_baseline)

# reencode the sex and APOE4 carrier status
mci_baseline = mci_baseline %>%
  mutate(sex = if_else(sex == 0, 'female', 'male')) %>%
  mutate(apoe4 = if_else(apoe4 == 0, 'non-carrier', 'carrier'))

# remove the participants who don't meet the stated inclusion criteria
mci_base_new = mci_baseline %>%
  mutate(age_at_onset = ifelse(age_at_onset == '.', NA, age_at_onset)) %>%
  filter(current_age < age_at_onset | is.na(age_at_onset))

```

**Discuss important steps in the import process and relevant features of the dataset.**

The important steps in the import process is that we need to skip the first row since it's notes that we don't need in our dataset otherwise it would cause messy. But the note gives us a basic understanding about how the origin data were encoded. For example, Sex == 0 represents the participant is female. apoe4 == 0 represents the participant is non-carrier. We need also focus on `age at onset` variable since we need use it to filter out the participants who don't meet the criteria.

**How many participants were recruited, and of these how many develop MCI?**

`r nrow(mci_baseline)` participants were recruited. `r nrow(mci_base_new)` develop MCI.

**What is the average baseline age?**

```{r}
# calculate for the average baseline age
base_mean_age_mci = mean(pull(mci_base_new, current_age))

```

For those develop MCI, the average baseline age is `r mean(pull(mci_base_new, current_age))`.

**What proportion of women in the study are APOE4 carriers?**

```{r}
# calculate the proportion
num_carrier_female = mci_base_new %>%
  filter(sex == 'female') %>%
  filter(apoe4 == 'carrier') %>%
  count()

num_female = mci_base_new %>%
  filter(sex == 'female') %>%
  count()

pro_female_carrier = num_carrier_female/num_female
```

The proportion of women in the study who are APOE4 carriers is `r pro_female_carrier` .

**Similarly, import, clean, and tidy the dataset of longitudinally observed biomarker values; comment on the steps on the import process and the features of the dataset.**

```{r}
# load the amyloid dataset
mci_amyloid = read.csv('./dataset/mci_amyloid.csv', skip = 1)

# clean the data
mci_amyloid = janitor::clean_names(mci_amyloid)

# change the name of variables in the dataset
mci_amyloid = mci_amyloid %>%
  rename(id=study_id,
         year_0 = baseline,
         year_2 = time_2,
         year_4 = time_4,
         year_6 = time_6,
         year_8 = time_8) %>%
  pivot_longer(
    year_0:year_8,
    names_to = "time",
    values_to = "ratio"
  )
  

```

For this dataset, we also need to skip the first row since it's also note. We rename `study_id` to `id` for later combination and we rename `baseline` to `year_0` for consitency. We apply pivot_longer( ) to make the dataset more readable with time and ratio obviously shown. There are some NAs in the dateset, which means some ratio values are missing. It totally has `r nrow(mci_amyloid)` observations (where an id has 5 rows, which means `r nrow(mci_amyloid)/5` participants) and `r ncol(mci_amyloid)` variables. It contains key variables such as id, time and ratio.

**Check whether some participants appear in only the baseline or amyloid datasets, and comment on your findings.**

```{r}
baseline_only = anti_join(mci_base_new, mci_amyloid,by = "id")
amyloid_only = anti_join(mci_amyloid, mci_base_new, by = "id")

# find the difference id in each dataset
only_in_baseline_id = setdiff(mci_base_new$id, mci_amyloid$id)
only_in_amyloid_id = setdiff(mci_amyloid$id, mci_base_new$id)
```

There are totally `r nrow(baseline_only)` participants only appear in baseline dataset : participants with id `r only_in_baseline_id`.

There are totally `r nrow(amyloid_only)/5` participants only appear in amyloid dataset : participants with id `r only_in_amyloid_id`.

**Combine the demographic and biomarker datasets so that only participants who appear in both datasets are retained, and briefly describe the resulting dataset; export the result as a CSV to your data directory.**

```{r}
# combine two datasets together
combined_dataset = inner_join(mci_base_new, mci_amyloid, by = "id")
```

There are `r nrow(combined_dataset)/5` participants appear in both dataset, each participants have 5 rows data (year 0 - year 8). The result table contains the information of the participant's id, age, sex, education, apoe4 status, MCI development, change based on time.

```{r}
# export the result
write.csv(combined_dataset, file = "./dataset/combined dataset.csv", row.names = T)
```
